The Evolution of Phenotypically Invariant Genetic Networks

### 2017-06-20 handnotes.tex
conceptual figure

\begin{abstract}
 % \textbf{
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space. Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function. 
%}
\end{abstract}
\section*{Introduction}

  \jss{add stuff on Haldane's Rule and Hybrid Vigor; note: F1s diverge to the 4th power, and F2s quadratically}

@@

@@

  At any given time, there will be a range of network coefficients present in the population
  due to segregating genetic polymorphism.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?

  The amount and structure of 
  this standing variation is established over long time scales
  by many factors, including
  mutation-selection balance, 
  shifts in the phenotypic optimum,
  and/or spatial variation in the optimum \citep{sources_of_standing_var}.
  Quantitative genetics models of mutation-selection balance 
  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel},
  but it is unclear how well these predictions match reality \citep{johnson_barton}
  and how much they are expected to change over time \citep{arnold_changing_G}.
  However, recent empirical work allows us to estimate at least the rough magnitude of variation.
  XXX FILL IN DETAILS: 
  Mutational scans find this distribution of binding site change per nucleotide change in promoters.
  Intraspecific variation is often in the range of 0.1\%--1\%.
  This suggests typical regulatory binding strengths within a population might vary by up to roughly X\%.
  But, does it?
  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila,
  which suggests that (a) the observed range of variation is somewhat less than this,
  and (b) the strength of selection on phenotype is sufficient to weakly constrain this variation.

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Erik Lundgren, and Hossein Asgharian for valuable discussion.

@@

For completeness, we provide a brief argument of how the population mean
moves under genetic drift
with a quantitative genetics model.

@@

\plr{In case this is useful...}

@@

By section \ref{ss:quant_gen},
if we set $\Sigma=\sigma^2 I$ and $U=H$,
then a population at $A_0+Z$ experiences a restoring force of strength
$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
If $\sigma^2$ is small compared to $H^{-1}$
then this is approximately $-\sigma^2 H^{-1} Z$.

>>>>

\begin{abstract}
 % \textbf{
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space. Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function. 
%}
\end{abstract}

Additional ideas to consider adding:
\begin{itemize}
    \item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    \item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    \item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    \item obtain estimates of variation in $A$ from thermodynamic occupancy model
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

@@

  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}

@@

  At any given time, there will be a range of network coefficients present in the population
  due to segregating genetic polymorphism.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?
  To answer these questions, we need to know two things:
  the strength of stabilizing selection on the phenotype,
  and the amount (and structure) of heritable variation in the genotype.
  Both are hard to measure, somewhat ambiguous, and have values that may depend strongly on the context.
  We aim to get order-of-magnitude estimates.

  Specifically, we want to know how much the entries of $A$ vary within a population.
  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
  per change in concentration of $j$.
  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
  In the context of transcription factor networks this may be affected 
  not only by the binding strength of $j$ to the promoter region of $i$
  but also the effects of other transcription factors (e.g., cooperativity)
  and local chromatin accessibliity \citep{stefflova2013cooperativity}.
  (Recall that although these are best modeled through nonlinear terms,
  by linearizing we essentially consider first-order effects.)
  Furthermore, the mutational target size for variation in $A$ may be much larger than the kilobase or so of a typical promoter sequence.
  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
  in the presence of substantial molecular noise \citep{canalization,buffering},
  suggesting that substantial variation in the micro-scale dynamics we consider here
  may be necessary to produce relevant phenotypic effects downstream.
  \plr{replace "micro-scale" with sthg else or discuss earlier}

  The amount and structure of this standing variation is established over long time scales
  by many factors, including
  mutation-selection balance, 
  shifts in the phenotypic optimum,
  and/or spatial variation in the optimum \citep{hansen1996translating}.
  Quantitative genetics models of mutation-selection balance 
  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
  but it is unclear how well these predictions match reality \citep{johnson_barton}
  and how much they are expected to change over time \citep{arnold_changing_G}.
  However, empirical work allows us to estimate at least the rough magnitude of variation.
  XXX FILL IN DETAILS: 
  Mutational scans find this distribution of binding site change per nucleotide change in promoters.
  Intraspecific variation is often in the range of 0.1\%--1\%.
  This suggests typical regulatory binding strengths within a population might vary by up to roughly X\%.
  But, does it?
  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila,
  which suggests that (a) the observed range of variation is somewhat less than this,
  and (b) the strength of selection on phenotype is sufficient to weakly constrain this variation.

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.

@@

For completeness, we provide a brief argument of how the population mean
moves under genetic drift
with a quantitative genetics model,
as in \citet{lande1981models} or \citet{hansen1996translating}.
These ignore details of the underlying genetic basis,
but developing a more accurate model is beyond the scope of this paper.

@@

@@

By section \ref{ss:quant_gen},
if we set $\Sigma=\sigma^2 I$ and $U=H$,
then a population at $A_0+Z$ experiences a restoring force of strength
$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
If $\sigma^2$ is small compared to $H^{-1}$
then this is approximately $-\sigma^2 H^{-1} Z$.
This suggests that the population mean follows an Ornstein-Uhlenbeck process,
as described (in different terms) in \citet{hansen1996translating}.

### 2017-06-20 promoterdata.tex
conceptual figure

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\usepackage{multicol}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\newcommand{\calA}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

@@

    \begin{itemize}
      \item \citet{gertz2009analysis} study synthetic yeast promoters. The synthetic promoters contained a random combination of 3-4 TFBSs. Weak MIG1 TFBS has a 6.7-9.0 fold lower affinity than the strong TFBS. Weak sites in concert with a strong site act as a strong site too.

>>>>

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)

@@

\section{Nonlinearity in TF networks}

A common model is to use the Hill equation:
if $j$ upregulates $i$ then
\begin{align}
    \dot x_i(t) = \beta \frac{x_j(t)^n}{ K_A^n + x_j(t)^n}
\end{align}
or if it downregulates,
\begin{align}
    \dot x_i(t) = \beta (1-\frac{x_j(t)^n}{ K_A^n + x_j(t)^n} ).
\end{align}
Citation: Alon's book? gertz2008analysis?

@@

\begin{itemize}

    \item \citet{kasowski2010variation} use ChIP-seq to survey variation in binding sites between 10 human lymphoblastoid cell lines:
        RNA PolII (not a TF like we are considering) and NFkB (p65, a regulator of immune responses, which does have binding motifs);
        individuals differed at 25\% and 7.5\% of the sites respectively. 
        Also found 32\% binding sites having significant difference in PolII occupancy between human/chimp
        \begin{itemize}
            \item Differences explained by SNPs, structural variants, and other things (like \citet{shi2016evaluating})
            \item ``Significant differences in binding'' have SD of +/- 1 log2 difference 
            \item Weak connection of binding differences to expression differences:
                ``We examined the effect of binding variation on gene expression by generation of deep RNASeq
                data from each cell line (9) and comparison with binding data (Fig. 3A, Fig. S9A). A
                significant correlation was observed (Spearman correlation coefficients of 0.475 and 0.461 for
                NFκB and PolII, respectively) (Fig. 3B, Fig. S9B, Table S9), suggesting an influence of binding
                differences on mRNA abundance''
            \item "Since humans and chimpanzees exhibit 5–10\% differences in gene expression (15)"
        \end{itemize}

    \item \citet{schmidt2010vertebrate} use ChIP-seq to compare TF occupancy (for CEBPA and HNF4A) between five vertebrates (human, mouse, dog, opossum, and chicken), 
        and find (a) conservation of binding motif; (b) little sharing of actual binding sites; and (c) a fair amount of compensatory changes
        \begin{itemize}

            \item "Among the binding events lost in one lineage, only half are recovered by another binding event within 10 kilobases"

            \item "CEBPA and HNF4A were selected as representative transcription factors within the liver-specific regulatory
                network because both are conserved and constitutively expressed with well-characterized target genes (10, 11). In addition,
                they represent distinct TF classes, and the DNA binding domains of each factor's orthologs are nearly identical among the study species "

            \item "About 250 direct functional HNF4A target genes have recently been identified"

            \item both have consensus motifs of about 5 of 10 bp

            \item "only between 6-8 \% of the genomic regions occupied by CEBPA in opossum liver align with CEBPA binding events also found in
                mouse, dog, and/or human liver. This divergence was even greater in chicken, which shared only 2\% of CEBPA binding with human"
        \end{itemize}

    \item \citet{shi2016evaluating} looked at variation in TF binding in heterozygotes in ENCODE data
        and found that differential binding is often explained by SNPs in the TF's motif (19.3\%) or potential partner TFs' motifs (9.4\%);
        variation in chromatin accessibility; or histone modifications (NOT just SNPs).
        \begin{itemize}
            \item shifts of motif score of 0.05--0.1 were responsible for differential occupancy
            \item "While altered canonical TFBSs for the ChIP'd TFs were frequently observed (19.3\%), most ASB SNVs did not overlap with the primary TF motif."
        \end{itemize}

    \item \citet{stefflova2013cooperativity} In five Mus (sub-)species, found that
        ``Although individual mutations in bound sequence motifs can influence TF binding, most binding differences occur in the absence of nearby sequence variations.''
        \begin{itemize}
            \item "Clustered TF binding appears to result in large part
                from indirect cooperativity to open chromatin regions, as opposed
                to direct TF-TF protein interactions (Kaplan et al., 2011; Miller
                and Widom, 2003; Mirny, 2010). For binding sites within a
                nucleosome-length distance, each TF contributes partially to a
                competitive displacement of specific nucleosomes by indirect
                collaboration with other TFs, mutually aiding each others’ binding
                to DNA. TFs within a cluster can have different regulatory roles
                depending on their motif strength and ability to compete with
                nucleosomes (Zinzen et al., 2009)." 
        \end{itemize}

    \item \citet{tugrul2015dyanmics} do theory on the gain and loss of TFBS.  
        Conclusion: more than 10bp unlikely; de novo very hard but made possible ad the scale of eukaryotic species divergence by pre-existing relics, cooperativity, etc.
        Euk TFBS typically 6--12bp.

  \item \citet{gertz2009analysis} study synthetic yeast promoters. The synthetic promoters contained a random combination of 3-4 TFBSs. Weak MIG1 TFBS has a 6.7-9.0 fold lower affinity than the strong TFBS. Weak sites in concert with a strong site act as a strong site too.
